---
title: "Lecture 8"
output: pdf_document
date: "2024-09-30"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ps270data)
```

## Observational Studies

Do newspaper endorsements matter?
  Can newspaper endorsements change voters' minds?
  Why not compare vote choice of readers of different papers? 
    Problem: readers choose papers based on their previous beliefs 

Study for today: British newspapers switching their endorsements
  Some newspapers endorsing Tories in 1992 switched to Labour in 1997.
  Treated group: readers of Tory -> Labour papers.
  Control group: readers of Tory who didn't switch. 
  
Load dataset:

```{r}
newspapers
```
  
Example of an observational study:
  We as researchers observe a naturally assigned treatment.
  Very common: often can't randomize for ethical/logical reasons. 
  
Internal validity: are the causal assumption satisfied? Can we interpret this as a causal effect? 
  RCTs usually have higher internal validity
  Observational studies less so because treatment and control groups may differ in ways that are hard to measure 

External validity: can the conclusions/estimated effects be generalized beyond this study?
  RCTs weaker here because often very expensive to conduct on representative samples
  Observational studies often have larger/more representative samples that improve external validity
  
Confounder: pre-treatment variable affecting treatment & the outcome. 

Confounding bias in the estimated SATE fue to these differences. 
  If our control group is different from our treatment group in terms of confounders, then it's not a good proxy
  
## Research designs

How can we find a good comparison group?

Depends on the data we have available.

Three general types of observational study research designs:
1. Cross-sectional design: compare outcomes treated and control units at one point in time.
      Assumption: groups identical on average; sometimes called unconfoundedness or as-if randomized 
2. Before and after design: compare outcomes before and after a unit has been treated, but need over-time data on treated group
      Advantage: all person-specific features held fixed
      Assumption: no time-varying confounders
3. Difference-in-differences design: use the before/after difference of control group to infer what would have happened to treatment group without treatment 
      Change in treated group above and beyond the change in control group
      Assumption: parallel trends 
  
Let's calculate the cross-sectional estimate:

```{r}
switched <- newspapers |>
  filter(to_labour == 1) |>
  summarize(mean(vote_lab_97))

no_change <- newspapers |>
  filter(to_labour == 0) |>
  summarize(mean(vote_lab_97))

switched - no_change
```
  
Statistical control: adjust for confounders using statistical procedures
  Can help to reduce confounding bias
  
One type of statistical control: subclassification
  Compare treated and control groups within levels of a confounder
  Remaining effect can't be due to the confounder 
  
Threat to inference: we can only control for observed variable -> treat of unmeasured confounding 

Statistical control on R:

```{r}
newspapers |>
  group_by(parent_labour, to_labour) |>
  summarize(avg_vote = mean(vote_lab_97)) |>
  pivot_wider(
    names_from = to_labour,
    values_from = avg_vote) |>
  mutate(diff_by_parent = `1` - `0`)
```
  
Before and after in R:

```{r}
newspapers |>
  filter(to_labour == 1) |>
  mutate(
    vote_change = vote_lab_97 - vote_lab_92) |>
  summarize(avg_change = mean(vote_change))
```
  
Differences-in-differences in R:

```{r}
newspapers |>
  mutate(
    vote_change = vote_lab_97 - vote_lab_92,
    to_labour = if_else(to_labour == 1, "switched", "unswitched")
  ) |>
  group_by(to_labour) |>
  summarize(avg_change = mean(vote_change)) |>
  pivot_wider(
    names_from = to_labour, 
    values_from = avg_change
  ) |>
  mutate(DID = switched - unswitched)
```
  
  
  
  
  
  