---
title: "Lecture 12"
output: pdf_document
date: "2024-10-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tidying and Joining Data 

## Pivot Longer

First let's load our packages:

```{r}
library(tidyverse)
library(ps270data)
mortality
```

to convert a data set into the long format, use the pivot_longer() function

mydata |>
  pivot_longer(
    cols = <variables to pivot>,
    names_to = <new vairable to put column names>,
    values_to = <new variable to put column values>
)

Let's do it with the mortality data

```{r}
mortality |>
  select(-indicator) |>
  pivot_longer(
    cols = `1972`:`2020`,
    names_to = "year",
    values_to = "child_mortality"
  )
```

let's do a line plot

```{r}
mortality |>
  select(-indicator) |>
  pivot_longer(
    cols = `1972`:`2020`,
    names_to = "year",
    values_to = "child_mortality"
  )  |>
  mutate(year = as.integer(year)) |>
  ggplot(mapping = aes(x = year, y = child_mortality, group = country)) +
  geom_line(alpha = 0.25)
```

let's practice pivot_longer on another dataset

```{r}
spotify
```

```{r}
spotify |>
  pivot_longer(cols = c(-`Track Name`, -`Artist`),
               names_to = "week_of_year",
               values_to = "rank",
               names_prefix = "week") |>
  mutate(week_of_year = as.integer(week_of_year))

```

## Joining Data Sets

```{r}
library(gapminder)
gapminder
```

first, assign our pivoted mortality data to the object mortality_long

```{r}
mortality_long <- mortality |>
  select(-indicator) |>
  pivot_longer(
    cols = `1972`:`2020`,
    names_to = "year",
    values_to = "child_mortality"
  )  |>
  mutate(year = as.integer(year))
```

```{r}
mortality_long
```

Check that keys are unique

```{r}
gapminder |>
  count(country, year) |>
  filter(n > 1)
```

same for the other data

```{r}
mortality_long |>
  count(country, year) |>
  filter(n > 1)
```

first we use the left_join() function

```{r}
gapminder |>
  left_join(mortality_long)
```

an alternative (that does something different) is inner_join()

```{r}
gapminder |>
  inner_join(mortality_long)
```

















